Optimally Efficient Collaborative Text Editing

### 2012-12-14 paper-collabed.tex
thorough revision of paper

\subsection{The Monotone Shared Edit Tree (MSET) approach}
In this paper we present a new algorithm based on a more sophisticated
representation of the string. We call this the 
{\bf Monotone Shared Edit Tree} or MSET representation which is based 
on a model, view, controller paradigm.  The key idea is to use a model for
string editing that maintains three simultaneous views: the usual string editing view, the "tracking changes" view which shows all of the previously "deleted" code highlighted in someway (e.g. with strikethrough font), and an "edit-string" view which gives a more full record of the editing session recording which users made which insertions among other information. Thus, the editing session will be implemented by maintaining a set of models $\{M^u_i: i=0,1,2,\ldots\}$ for each user $u$. These models have several views including the standard view which is the only view in which the user is allowed to edit.  Two local editing operations are allowed: 
\begin{itemize}
\item insert a character $\alpha$ at offset $k$
\item delete the character $\alpha$ at offset $k$
\end{itemize}
and each of these is transformed into an operation $\tau$ on the underlying model and that operation is broadcast to all other users editing the document. Thus,
if we let $\alpha^u_i$ denote the standard view of the string that user $u$ is editing after $i$ edit steps have been applied and if we let $\sigma^u_i$ denote then the
editing operation applied on the string and $\tau^u_i$ denote the corresponding MSET operation, then the following diagram commutes:
\[
\begin{diagram}
\node{M^u_i} \arrow{e,t}{\tau^u_i} \arrow{s,l}{\gamma} 
\node{M^u_{i+1}} \arrow{s,r}{\gamma} \\
\node{\alpha^u_i} \arrow{e,t}{\sigma^u_i} \node{\alpha^u_{i+1}}
\end{diagram}
\]
where $\gamma$ is the projection that converts an MSET model $M$ into the corresponding string view $\alpha=\gamma(M)$.  When user $u$ performs a local edit $\sigma^u_i$ the system converts that into an operation $\tau^u_i$ on the model, and that operation is applied to the underlying model (thereby changing the standard string view) and is simultaneously broadcast to all peers.  Likewise, when the system receives an MSET transformation from a peer, it places it on a queue and when the user is not editing, it dequeues a transformation $\tau^u_i$ to the model. This corresponds to a string edit operation $\sigma^u_i$ which is what the user sees.  The point is that the MSET system works both ways: transforming local string edits into MSET edits, and transforming remote MSET edits into local string edits.

@@

Let $u$ be the user performing one of the following tree tree edit operations
on the target node {\tt v/m:p}.
\begin{itemize}
\item ${\tt treeextend}(v/m:p,\alpha)\;\;\text{if u=v}$ \newline
First, user $u$ can extend the text in the node $v/m$
provided it is the owner (that is $v=u$). It extends the node by appending
some text $\alpha$ to the location $p$ at the end of the node,
Note however, that users can not modify
the text in any node that they do not own and can not change any of the text
once it has been put into the node.
\item ${\tt treeinsert}(v/m:p,q,u/n,\alpha) \;\; \text{if $q\le p$}$
\newline
Second, user $u$ can insert a new node $u/n$ containing a string $\alpha$
at any position $q\le p$ of the node, 
$v/m$ provided
they have not already inserted a node there.
\item ${\tt treehide}(v/m:p,q)  \;\; \text{if $q\le p$}$\newline
Third, user $u$ can set the visibility of a character at position $q < p$
in  $v/m:p$ to false.
When characters are initially added to a node they have visibilty true, but once
set to false, the visibilty must remain false. 
\end{itemize}
For all of these operations, we say that the node $v/m:p$ is the target of the edit operation.

Given these rules we see that a group of distributed users can easily edit a 
shared tree by keeping their own representation of the tree and broadcasting
their edits to their peers. Observe that these edit operations commute in
the following sense. The only place where two user's could have a editing
conflict is when they both want to insert a string of characters at a particular
position $q$ in some node $v/m:p$, but in this case their insertions are always
ordered by userid and this resolves the conflict. No matter which order the
operations are applied the same tree results. This is phrased more formally
in the following lemma:

\begin{lemma}
Let $T$ be any edit-tree and let $\tau_1,\tau_2,\ldots,\tau_n$
be a sequence of valid edit operations on $T$ as described above.
This yields a sequence of edit trees $T_1,\ldots,T_n$. Let 
$\sigma_1,\ldots,\sigma_n$ be any reordering of this sequence of operations
with the property that if $\sigma_i$ has target $v/m:p$, then $v/m:p$ either
exists in $T=T_0$, or $v/m:p$ was created by one of the preceeding $\sigma_i$
which must either be an insert or an extend operation.
Under these assumptions, the corresponding sequence of trees
$S_1,\ldots,S_n$ is well-defined and $S_n=T_n$.
\end{lemma}

@@

\begin{lemma}
Let $G$ denote a collaborative editing session as described
above and assume that all clients $C_1,\ldots,C_n$ follow the protocol above
for a certain period of time and then stop. Once all queues $Q_i$ have been
emptied and all operations processed, all users will have the same edit
trees, that is $T_1=T_2=\ldots=T_n$.
\end{lemma}

@@

\section{The unoptimized model for edit trees with a edit-string view}
We are really interested in collaborative editing of strings, not trees, so the
question arises as to how we can convert string edit operations of the user
into tree edit operations on the user's model, and how we can convert tree edit
operations from the user's peers into string edit operations on the user's string,
and how can we efficiently broadcast these edit operations to all other members
of the collaborative edit session.

As a first step, observe that there is a one-to-one correspondence between
edit-strings as defined above and edit-trees, thus our distributed algorithm
for edit-trees provides a distributed algorithm for edit-string where we
assume that the clients $C_i$ each maintain a edit-string $S_i$ and perform valid
edit-string operations on $S_i$. These operations can be transformed directly
into valid tree operations which can be broadcast to the peers above.

@@

We describe the valid edit-string insertion operations as insertions
between two symbols $d_i$ and $d_{i+1}$ in the edit-string, where the $d_i$
is one of the following:
\begin{itemize}
\item  a visible character (denoted $c_i$),
\item a hidden character (also denoted $c_i$),
\item  a start marker $<^v_m$ or
\item an end marker $>^v_m$
\end{itemize}
Let $u$ denote the user performing
the operation and let $n$ be the next available block id for user $u$, so if
the user inserts a new block containing the string $\alpha$, it will be inserted
as $<^u_n \alpha >^u_n$.  Let $c_1$ and $c_2$ denote non-marker characters
which could be either visible or hidden or one of each and let
 $v$ and $w$ be users different from $u$ and each other.

\begin{lemma} Notation as above. The following are precisely the set of valid insert operations, where we indicate the two elements on either side of the insert on the left and the result on the right, together with the condition (if any) that must be satisfied for that insertion to be allowed:
\begin{eqnarray}
c_1\;\;\;\;\; c_2 &\rightarrow& c_1 <^u_n \alpha >^u_n c_2 \label{eq:ins1}\\
c_1\;\; >^w_m &\rightarrow& \left\{ 
  \begin{array}{l l}
      c_1 \alpha >^w_m &  {\rm if}\; u=w  \\
      c_1 <^u_n \alpha >^u_n >^w_m &  {\rm if} \; u \ne w 
  \end{array}  \right . \label{eq:ins2}\\
c_1 \;\;<^w_m &\rightarrow& \left\{ 
\begin{array}{l l}
c_1 <^u_n \alpha >^u_n <^w_m & {\rm if}\; u<w  \\
FAIL & {\rm if} u \ge w \\
\end{array}
\right . \label{eq:ins3}\\
>^v_r \;\;c_2 &\rightarrow& 
 \left\{ 
  \begin{array}{l l}
    >^v_r  <^u_n \alpha >^u_n c_2 & {\rm if} \; v<u \\
    FAIL & if v \ge u \\
  \end{array} \right .\label{eq:ins4}\\
>^v_r \;\;>^w_m &\rightarrow& 
  \left\{
    \begin{array}{l l}
      >^v_r \alpha >^w_m &{\rm if}\; u=w \\
      >^v_r <^u_n \alpha >^u_n >^w_m & {\rm if} \; (v<u)  \&  (v \ne w) \\
     FAIL & {\rm (v > u) \& (v \ne w)}
    \end{array}\right . \label{eq:ins5} \\
>^v_r \;\;<^w_m &\rightarrow & 
\left \{
\begin{array}{l l}
 >^v_r <^u_n \alpha >^u_n <^w_m & {\rm if} \; v<u<w \\
FAIL & {\rm otherwise} 
\end{array} \right . \label{eq:ins6}\\
<^v_r \;\;c_2 &\rightarrow& <^v_r  <^u_n \alpha >^u_n c_2\label{eq:ins7}\\
<^v_r \;\;>^w_m &\rightarrow& 
\left \{
\begin{array}{l l}
<^v_r <^u_n \alpha >^u_n >^w_m & {\rm if} \; v=w=0, r=m=1\\
FAIL & {\rm otherwise} \\
\end{array} \right . \label{eq:ins8}\\
<^v_r \;\;<^w_m &\rightarrow&
\left \{
\begin{array}{l l}
 <^v_r <^u_n \alpha >^u_n <^w_m \;\;\; {\rm if} \; u<w \\
FAIL & {\rm otherwise} \\
\end{array} \right . \label{eq:ins9}
\end{eqnarray}
\end{lemma}

Note that user $u$ can always insert a new node between any two non-marker
characters (Eq. \ref{eq:ins1}), and user $u$ can always insert a string
directly into the end of a node that it owns ($u=w$ cases in Eqs \ref{eq:ins2}, \ref{eq:ins5}).
Also, the only empty node is the root of the tree $<^0_1\;\; >^0_1$ and this
is the only time you have a start marker directly followed by a end marker
(Eq. \ref{eq:ins8}).  The final case to take note of is Eq. \ref{eq:ins7} where
$u$ inserts a string at the between a start-marker and a non-marker characters.
We will come back to these five insertion cases in the next section.

For now, we observe that if a edit-string view only allows the insertions
as shown above for user $u$, then each such insertion corresponds to a valid
edit-tree operation and hence the collaborative editing protocol in the previous
section for edit-trees can be naturally extended to edit-strings. The following
lemma states the convergence property for this distributed algorithm.

@@

This model is only slightly more complicated than the previous case of
collaborative editing of edit-strings.  In this case, each client maintains
both a standard view of a string (with no deletions and no start/end markers)
and they also maintain the edit-string and edit-tree models in such a way
that these models all agree.

Our first observation is that each character in the standard view corresponds
to a uniquely defined visible non-marker character in the edit-string view. 
Given that observation,
we need to consider the two string operations that a user can perform on a
standard view string and see how we can transform that to an operation on 
the edit-string view.

There are two operations we allow on the standard view:
\begin{itemize}
\item ${\tt insert}(p,\alpha)$ - insert the string $\alpha$ at offset $p$ in the string
\item ${\tt remove}(p,\alpha)$ - remove the substring $\alpha$ begining at offest $p$
which has the effect of changing the attributes of those characters from visible to hidden.
\end{itemize}

If $\gamma$ is a edit-tree and $f(\gamma)$ denotes the standard string view
of that tree, then we want to show that for each string operation $\sigma$
on $f(\gamma)$ there is a corresponding valid operation $\tau$
on the edit-tree, such that 
\[
f(\tau(\gamma)) = \sigma(f(\gamma))
\]
This will allow us to convert string operations into edit-tree operations which
can then be broadcast to all user as in the previous section. There may be many possible $\tau$ operations that satisfy this property (and often there are many), but all we need is the existence of at least one such operation. At this point, we don't worry about the computation complexity of calculating the edit-tree operation corresponding to a string operation, that will be the subject of the next section.

The remove operation is easily transformed into a edit-tree hide operation which 
simply sets the visibility attribute of the corresponding character to ``hidden''.

@@

There are four basic insert operations on the string $f(\gamma)$ in the standard view to consider.
\begin{itemize}
\item inserting into an empty string
\item inserting at the beginning of the string
\item inserting at the end of the string
\item inserting somewhere in the middle of the string 
\end{itemize}
The first case is easily handled as it results in the following operation on the 
edit-string $\gamma$:
\[
<^0_0\;>^0_0 \;\; \rightarrow \;\;<^0_0 <^u_0 \alpha >^u_0 >^0_0
\]
The second is a little trickier. We need to find the first non-marker character $d_k$
in the full-edit string and observe that the previous character must be a start
marker $<^v_m$ so we can insert the new node $u/n$ containing $\alpha$ between those
two points, where $n$ is the number of nodes owned by $u$ in $\gamma$ and observe
that inserting a new node 
\[
<^u_n \alpha >^u_n
\]
between a start marker and a non-marker character is always valid.

@@

The last case is when we are inserting between two characters $c_i$ and $c{_i+1}$
in the standard view $f(\gamma)$. Let $d_j$ denote the visible non-marker
character corresponding to $c_i$ in the edit-string $\gamma$ and let
$d_k$ denote the next non-marker character. So $d_j$ and $d_k$ are non-marker
characters, all the symbols between them are marker symbols, $d_j$ is
visible, and $d_k$ could be visible or hidden. There are three subcases:
\begin{itemize} 
\item $d_{j+1}$ is a non-marker character, i.e. $d_j$ and $d_k$ are adjacent. 
In this case, inserting
the node $u/n$ between them is always valid.
\item $d_{j+1}$ is an end marker $>^v_m$, again in this case inserting $u/n$
between $d_j$ and $d_{j+1}$ will be a valid insert operation.
\item $d_{j+1}$ is a start-marker. In this case all of the symbols between
$d_j$ and $d_k$ must be start markers, so $d_{k-1} = <^v_m$ is a start marker
and inserting between that symbol and the non-marker character that follows
is a valid operation.
\end{itemize}
In each of these three cases, we get a edit-string operation $\tau$ that
corresponds to inserting $\alpha$ between $c_i$ and $c_{i+1}$. The arguments
above have proven the following lemma.

\begin{lemma}
Let $\gamma$ be a edit-tree and let $f(\gamma)$ be its standard string view, then
the algorithm described above constructs a valid edit-tree operation $\tau$
on $\gamma$ for each valid edit operation $\sigma$ on the string $f(\gamma)$, and
we have
\[
\sigma(f(\gamma)) = f(\tau(\gamma))
\]
\end{lemma}

We can now define a
collaborative editing algorithm where the $i$th client presents 
a standard view of a
string $\alpha_i$
to the $i$th user who can then perform arbitrary string
operations on that string.

These string operations $\sigma$ on $\alpha_i$ 
are then transformed into valid edit-tree operations $\tau$
on the underlying edit-tree $T_i$ and these operations are then handled as in the
collaborative edit tree operation. We again have a correctness proof. Moreover, 
this algorithm easily allows one to switch between the two  views: 
the standard string view and the edit-tree view (or more precisely the
edit-string view which is a string representation of the edit-tree).

@@

We have shown that edit-trees can be used to implement a collaborative editor
which allows the user to immediately apply local edits and which has the convergence
property.  The model as presented in the previous section becomes unacceptably slow as the size of the tree grows.  In this section, we show how to optimize this model by creating a more complex representation which allows
the string to tree and tree to string conversions on edit operations to be computed
in time $O(k\log(N))$ where $N$ is the size of the edit tree and $k$ is the number
of characters in the operation (i.e. number inserted or deleted).

In particular, the MSET algorithm will maintain the edit tree as described above,
but will also maintain an auxiliary data structure $M$ that allows it to rapidly
compute the node position in the edit tree corresponding to an element (i.e. a visible or hidden character or a start or end marker) in any one
of three string views, and vice versa. The three string views are
\begin{itemize}
\item $M.S$: the "standard" list of all visible characters
\item $M.T$: the "tracking changes" list of all visible and hidden characters
\item $M.E$: the "edit-string" list of all edit-string characters which faithfully represents the edit-tree by showing all visible and hidden characters and all markers (both
start markers and end markers)
\end{itemize}
We will show that these three lists can be implemented in such a way as to be able
to calculate the $k$th element of each list (e.g. M.E[k]) in time $O(\log(N))$ where $N$ is the size of the edit-tree.

>>>>

\subsection{The Monotone Shared Edit Tree (MSET) approach}
In this paper we present a new algorithm based on a more sophisticated
representation of the string. We call this the 
{\bf Monotone Shared Edit Tree} or MSET representation which is based 
on a model, view, controller paradigm.  The key idea is to use a model for
string editing that maintains three simultaneous views: the usual string editing view, the "tracking changes" view which shows all of the previously "deleted" code highlighted in someway (e.g. with strikethrough font), and an "edit-string" view which gives a more full record of the editing session recording which users made which insertions among other information. Thus, the editing session will be implemented by maintaining a set of models $\{T^u_i: i=0,1,2,\ldots\}$ for each user $u$. These models have several views including the standard view which is the only view in which the user is allowed to edit.  Two local editing operations are allowed: 
\begin{itemize}
\item insert a character $\alpha$ at offset $k$
\item delete the character $\alpha$ at offset $k$
\end{itemize}
and each of these is transformed into an operation $\tau$ on the underlying model and that operation is broadcast to all other users editing the document. Thus,
if we let $S^u_i$ denote the standard view of the string that user $u$ is editing after $i$ edit steps have been applied and if we let $\sigma^u_i$ denote then the
editing operation applied on the string and $\tau^u_i$ denote the corresponding MSET operation, then the following diagram commutes:
\[
\begin{diagram}
\node{T^u_i} \arrow{e,t}{\tau^u_i} \arrow{s,l}{\Gamma} 
\node{T^u_{i+1}} \arrow{s,r}{\Gamma} \\
\node{S^u_i} \arrow{e,t}{\sigma^u_i} \node{S^u_{i+1}}
\end{diagram}
\]
where $\Gamma$ is the projection that converts an MSET model $T$ into the corresponding string view $S=\Gamma(T)$.  When user $u$ performs a local edit $\sigma^u_i$ the system converts that into an operation $\tau^u_i$ on the model, and that operation is applied to the underlying model (thereby changing the standard string view) and is simultaneously broadcast to all peers.  Likewise, when the system receives an MSET transformation from a peer, it places it on a queue and when the user is not editing, it dequeues a transformation $\tau^u_i$ to the model. This corresponds to a string edit operation $\sigma^u_i$ which is what the user sees.  The point is that the MSET system works both ways: transforming local string edits into MSET edits, and transforming remote MSET edits into local string edits.

@@

Let $u$ be the user performing one of the following tree tree edit operations
on the target node {\tt v/m:p} to create a new node {\tt u/n:q}.
\begin{itemize}
\item ${\tt treeextend}(v/m:p,\alpha)\;\;\text{if u=v}$ \newline
First, user $u$ can extend the text in the node $v/m$
provided it is the owner (that is $v=u$). It extends the node by appending
some text $\alpha$ to the location $p$ at the end of the node,
Note however, that users can not modify
the text in any node that they do not own and can not change any of the text
once it has been put into the node.
\item ${\tt treeinsert}(v/m:p,q,u/n,\alpha) \;\; \text{if $q\le p$}$
\newline
Second, user $u$ can insert a new node $u/n$ containing a string $\alpha$
at any position $q\le p$ of the node, 
$v/m$ provided
they have not already inserted a node there.
\item ${\tt treehide}(v/m:p,q)  \;\; \text{if $q\le p$}$\newline
Third, user $u$ can set the visibility of a character at position $q < p$
in  $v/m:p$ to false.
When characters are initially added to a node they have visibilty true, but once
set to false, the visibilty must remain false. 
\end{itemize}
For all of these operations, we say that the node $(v/m:p)$ is the target of the edit operation and the node that is created, either $(v/m:p+1)$ for an extension or $(u/n:1)$ for a new node, is the result.

We use these definitions to define a partial order on edit operations $\tau\prec\tau'$ where $\tau$ directly preceds $\tau'$ if the result of $\tau$ is the target of $\tau'$, and we let $\prec$ be the transitive closure of that relation.  Then $\prec$ is a partial order on the set of all possible edit operations on a tree. We say that two operations $\tau$ and $\tau'$ are independent if neither precedes the other, that is 
\[
\left (\tau\not\prec\tau'\right ) \wedge  \left ( \tau'\not\prec\tau \right )
\]
Observe that operations performed simultaneously by clients on their own local edit trees are clearly independent!

Given these definitions and rules we see that a group of distributed users can easily edit a 
shared tree by keeping their own representation of the tree and broadcasting
their edits to their peers. Observe that these edit operations commute in
the following sense. The only place where two user's could have a editing
conflict is when they both want to insert a string of characters at a particular
position $q$ in some node $v/m:p$, but in this case their insertions are always
ordered by userid and this resolves the conflict. No matter which order the
operations are applied the same tree results. This is phrased more formally
in the following lemma:

\begin{lemma}
Let $T=T_0$ be any edit-tree and let $\tau_0,\tau_2,\ldots,\tau_{n-1}$
be a sequence of valid edit operations on $T$ as described above.
This yields a sequence of edit trees $T_0,\ldots,T_n$. Let 
$\tau'_0,\ldots,\tau'_{n-1}$ be any reordering of this sequence of operations
which is topologically sorted with respect to $\prec$, that is
with the property that if $i<j$ then either $\tau'_i\prec\tau'_j$ or
$\tau'_i$ and $\tau'_j$ are independent. Let $S_0=T_0$ and $S_{i+1} = \tau'_i(S_i)$.
Under these assumptions, the corresponding sequence of trees
$S_1,\ldots,S_n$ is well-defined and $S_n=T_n$.
\end{lemma}

@@

\begin{lemma}
Let $G$ denote a collaborative editing session as described
above and assume that all clients $C_1,\ldots,C_n$ follow the protocol above
for a certain period of time and then stop. Once all queues $Q_i$ have been
emptied and all operations processed, all users will have the same edit
trees, that is if $T_i$ is the final edit tree of $C_i$, then $T_1=T_2=\ldots=T_n$.
\end{lemma}

@@

\section{The unoptimized model for edit trees with a edit-string view}
We are really interested in collaborative editing of strings, not trees, so the
question arises as to how we can convert string edit operations of the user
into tree edit operations on the user's model, and how we can convert tree edit
operations from the user's peers into string edit operations on the user's string.

As a first step, observe that there is a one-to-one correspondence between
edit-strings as defined above and edit-trees, thus our distributed algorithm
for edit-trees provides a distributed algorithm for edit-strings where we
assume that the clients $C_i$ each maintain a edit-string $S_i$ and perform valid
edit-string operations on $S_i$. These operations can be transformed directly
into valid tree operations which can be broadcast to the peers above.

@@

We describe the valid edit-string insertion operations as insertions
between two symbols $d_i$ and $d_{i+1}$ in the edit-string, where the $d_i$
is one of the following:
\begin{itemize}
\item  a character, $c_i$, either visible or hidden,
\item  a start marker $<^v_m$ or
\item an end marker $>^v_m$
\end{itemize}
Based on this categorizations, there are nine possibilites for inserting a character between two adjacent elements in an edit string. In each case we could either enter a single character $\alpha$ or a node $<^u_n \alpha >^u_n$ containing that character. The next lemma considers all of those possibilities and specifies which ones correspond to legal operations on an edit tree.

Let $u$ denote the user performing
the operation and let $n$ be the next available block id for user $u$, so if
the user inserts a new block containing the character $\alpha$, it will be inserted
as $<^u_n \alpha >^u_n$.  Let $c_i$ and $c_j$ denote non-marker characters in
a block $v/m$ 
which could be either visible or hidden or one of each and let
 $v$ and $w$ be users different from $u$ and each other.

\begin{lemma} Notation as above. There are 9 cases for inserting a character or a node containing a character into an edit string between two elements $c_i$ and $c_{i+1}$ in an edit string within the scope of a node $v/m$. The following cases specify which possibilites correspond to valid operations on the edit tree and which do not. If a case is not listed, then it is not valid. 
\begin{eqnarray}
c_i\;\;\;\;\; c_{i+1} &\rightarrow& c_i <^u_n \alpha >^u_n c_{i+1} \label{eq:ins1}\\
c_i\;\; >^w_r &\rightarrow& \left\{ 
  \begin{array}{l l}
      c_i \alpha >^w_r &  {\rm if}\; u=w  \label{eq:ins2a}\\
      c_i <^u_n \alpha >^u_n >^w_r &  {\rm if} \; u \ne w 
  \end{array}  \right . \label{eq:ins2b}\\
c_i \;\;<^w_r &\rightarrow& \left\{ 
\begin{array}{l l}
c_i <^u_n \alpha >^u_n <^w_r & {\rm if}\; u<w  \label{eq:ins3a}\\
{\rm INVALID} & {\rm if} u \ge w \label{eq:ins3b}\\
\end{array}
\right . \label{eq:ins3}\\
>^v_m \;\;c_{i+1} &\rightarrow& 
 \left\{ 
  \begin{array}{l l}
    >^v_m  <^u_n \alpha >^u_n c_{i+1} & {\rm if} \; v<u \label{eq:ins4a}\\
   {\rm INVALID} & if v \ge u \label{eq:ins4b}\\
  \end{array} \right .\label{eq:ins4}\\
>^v_m \;\;>^w_r &\rightarrow& 
  \left\{
    \begin{array}{l l}
      >^v_m \alpha >^w_r &{\rm if}\; u=w \label{eq:ins5a} \\
      >^v_m <^u_n \alpha >^u_n >^w_r & {\rm if} \; (v<u)  \&  (v \ne w)  \label{eq:ins5b} \\
     {\rm INVALID} & {\rm (v > u) \& (v \ne w)} \label{eq:ins5c}
    \end{array}\right . \label{eq:ins5} \\
>^v_m \;\;<^w_r &\rightarrow & 
\left \{
\begin{array}{l l}
 >^v_m <^u_n \alpha >^u_n <^w_r & {\rm if} \; v<u<w  \label{eq:ins6a}\\
{\rm INVALID} & {\rm otherwise}  \label{eq:ins6b}
\end{array} \right . \label{eq:ins6}\\
<^v_m \;\;c_0 &\rightarrow& <^v_m  <^u_n \alpha >^u_n c_0\label{eq:ins7}\\
<^v_m \;\;>^w_r &\rightarrow& 
\left \{
\begin{array}{l l}
<^v_m <^u_n \alpha >^u_n >^w_r & {\rm if} \; v=w=0, r=m=0 \label{eq:ins8a}\\
{\rm INVALID} & {\rm otherwise} \label{eq:ins8b} \\
\end{array} \right . \label{eq:ins8}\\
<^v_m \;\;<^w_r &\rightarrow&
\left \{
\begin{array}{l l}
 <^v_m <^u_n \alpha >^u_n <^w_r & {\rm if} \; u<w  \label{eq:ins9a}\\
{\rm INVALID} & {\rm otherwise}  \label{eq:ins9b}\\
\end{array} \right . \label{eq:ins9}
\end{eqnarray}
\end{lemma}

\begin{proof}
Case (1): Note that user $u$ can always insert a new node between any two non-marker
characters (Eq. \ref{eq:ins1}), using operation $treeinsert(v/m:p,q,u/n,\alpha)$ but
since the position is not at the end of the node a straight insertion of $\alpha$ is not allowed.

Cases (2) and (5): User $u$ can always insert a string
directly into the end of a node that it owns via $treeextend(v/m:p,\alpha)$
 (Eqs. \ref{eq:ins2a}, \ref{eq:ins5a}) 
but the user can also insert a node at that position if they want $treeinsert(u/n:p,p,u/n,\alpha)$ if no node has been inserted there yet
 (Eq. \ref{eq:ins2b}). If another node has already been inserted at the
last position by user $v$ then user $u$ can only insert a node there if $v<u$.

Case (3):
A node $u/n$ can be inserted as the first node at an insertion point if the current
first node in the insertion point $w/m$ satisfies $u<w$.

Case (4):
A node $u/n$ can be placed at the end of an insertion point where the last node in the insertion is $v/m$ if and only if $v<u$.

Case (6):
A node $u/n$ can be inserted between two nodes $v/m$ and $w/r$ in an insertion point if and only if $v<u<w$.

Case (7): a node $u/n$ can always be placed at position 0 of a node $v/m$ provided nothing has yet been inserted there.

Case (8): 
The only case where a start symbol is immediately followed by an end symbol
if the root case $<^0_0>^0_0$ and in this case, user 0 is the root and does not insert any elements itself, so the only thing that can be inserted is a node, i.e. not a character (Eqs. \ref{eq:ins8a}, \ref{eq:ins8b} ).

Case (9): a node $u/n$ can be inserted at the position 0 of a node $v/m$ which already has some nodes inserted there if and only if the first node inserted a position 0, $w/r$ satisfies $u<w$.

\end{proof}

For now, we observe that if a edit-string view only allows the insertions
as shown above for user $u$, then each such insertion corresponds to a valid
edit-tree operation and hence the collaborative editing protocol in the previous
section for edit-trees can be naturally extended to edit-strings.  The converse
is also true, every valid edit-tree operation corresponds to one of the 9 valid cases above.

The following
lemma states the convergence property for this distributed algorithm.

@@

This model is only slightly more complicated than the previous case of
collaborative editing of edit-strings.  In this case, each client maintains three
views of the editing process: the standard view of the string currently being edits,
the "tracking changes" view which shows both the visible and "deleted" characters (with a different font of the deleted characters), and the edit-string view which
has characters (both visible and hidden with appropriate attributes for each type)
and markers (both start and end).

Our first observation is that each character in the standard view corresponds
to a uniquely defined visible character in the tracking-changes view, and
each character in the tracking-changes view corresponds to a non-marker character in the edit-string view. So we are really maintaining one list and two nested sublists of that list.

There are two operations we allow on the standard view:
\begin{itemize}
\item ${\tt insert}(p,\alpha)$ - insert the character $\alpha$ 
at offset $p$ in the string
\item ${\tt remove}(p,\alpha)$ - remove the character $\alpha$ at offest $p$
which has the effect of changing the attribute of that character from visible to hidden.
\end{itemize}

If $T$ is a edit-tree and $S = \Gamma(T)$ denotes the standard string view
of that tree, then we want to show that for each string operation $\sigma$
on $S$ there is a corresponding valid operation $\tau$
on the edit-tree, such that 
\[
\Gamma(\tau(T)) = \sigma(\Gamma(T))
\]
This will allow us to convert string operations into edit-tree operations which
can then be broadcast to all users as in the previous section. There can be more than one possible $\tau$ operation that satisfies this property (and often there are many), but all we need is the existence of at least one such operation. At this point, we don't worry about the computation complexity of calculating the edit-tree operation corresponding to a string operation, that will be the subject of the next section.

The remove operation is easily transformed into a edit-tree hide operation which 
simply sets the visibility attribute of the corresponding character to "hidden''.

@@

We now describe how to find one such $\tau$ and in the next section we present a data type in which that $\tau$ can always be constructed in time $O(\log(N))$.

There are four basic insert operations on the string $S=\Gamma(T)$ in the standard view to consider.
\begin{itemize}
\item inserting into an empty string
\item inserting at the beginning of the string
\item inserting at the end of the string
\item inserting somewhere in the middle of the string 
\end{itemize}
The first case is easily handled as it results in the following operation on the 
edit-string $\gamma$:
\[
<^0_0\;>^0_0 \;\; \rightarrow \;\;<^0_0 <^u_0 \alpha >^u_0 >^0_0
\]
The second is a little trickier. We need to find the first non-marker character $d_k$
in the full-edit string and observe that the previous character must be a start
marker $<^v_m$ so we can insert the new node $u/n$ containing $\alpha$ between those
two points, where $n$ is the number of nodes owned by $u$ in $T$ and observe
that inserting a new node 
\[
<^u_n \alpha >^u_n
\]
between a start marker and a non-marker character is always valid.

@@

The last case is when we are inserting between two characters $c_i$ and $c{_i+1}$
in the standard view $S = \Gamma(T)$. Let $d_j$ denote the visible non-marker
character corresponding to $c_i$ in the edit-string $T$ and let
$d_k$ denote the next non-marker character. So $d_j$ and $d_k$ are non-marker
characters, all the symbols between them are marker symbols, $d_j$ is
visible, and $d_k$ could be visible or hidden. There are three subcases:
\begin{itemize} 
\item $d_{j+1}$ is a non-marker character, i.e. $d_j$ and $d_k$ are adjacent. 
In this case, inserting
the node $u/n$ between them is always valid.
\item $d_{j+1}$ is an end marker $>^v_m$, again in this case inserting $u/n$
between $d_j$ and $d_{j+1}$ will be a valid insert operation.
\item $d_{j+1}$ is a start-marker. In this case all of the symbols between
$d_j$ and $d_k$ must be start markers, so $d_{k-1} = <^v_m$ is a start marker
and inserting between that symbol and the non-marker character that follows
is a valid operation.
\end{itemize}
In each of these three cases, we get a edit-string operation $\tau$ that
corresponds to inserting $\alpha$ between $c_i$ and $c_{i+1}$. The arguments
above have proven the following lemma.

\begin{lemma}
Let $T$ be a edit-tree and let $S=\Gamma(T)$ be its standard string view, then
the algorithm described above constructs a valid edit-tree operation $\tau$
on $T$ for each valid edit operation $\sigma$ on the string $S=\Gamma(T)$,
we have
\[
\sigma(\Gamma(T)) = \Gamma(\tau(T))
\]
\end{lemma}

We can now define a
collaborative editing algorithm where the user $u$ presents 
a standard view of a
string $S^u_i$ representing the effect of $i$ editing operations.
The user can then either insert or delete a character from that string
and broadcast the corresponding edit-tree operation to all peers.
Likewise, edit-tree operations that are received from peers can be applied to
the tree results in a modifed string $S^u_{i+1}$. Since the edit operations commute as long as they are independent, the order in which they are applied doesn't matter!
This proves the following theorem.

@@

We have shown that edit-trees can be used to implement a collaborative editor
which allows the user to immediately apply local edits and which has the convergence
property.  The model as presented in the previous section becomes unacceptably slow ($O(N)$ for each insert and delete) as the size of the tree grows if we use a naive data structure to represent the three lists and the edit tree.

In this section, we show how to optimize this model by creating a more complex representation which allows
the string to tree and tree to string conversions on edit operations to be computed
in time $O(\log(N))$ where $N$ is the size of the edit tree.

In particular, the MSET algorithm will maintain the edit tree $T$ as described above,
but will also maintain an auxiliary data structure $M$ that allows it to rapidly
compute the node position in the edit tree corresponding to an element (i.e. a visible or hidden character or a start or end marker) in any one
of three string views, and vice versa. The three string views are
\begin{itemize}
\item $M.S$: the "standard" list of all visible characters
\item $M.R$: the "tracking changes" list of all visible and hidden characters which shows all of the revision of the document,
\item $M.E$: the "edit-string" list of all edit-string characters which faithfully represents the edit-tree by showing all visible and hidden characters and all markers (both
start markers and end markers)
\end{itemize}
We will show that these three lists can be implemented in such a way as to be able
to calculate the $k$th element of each list (e.g. M.E[k]) in time $O(\log(N))$ where $N$ is the size of the edit-tree. Moreover, given an element $e$ its position in any one of these lists can also be computed in time $O(\log(N))$ using $M.pos(e,L)$ where $L$ is one of $S,R,E$.  In case, $e$ doesn't appear in a list of type $L$ (e.g. a marker character in the $S$ or $R$ list) we can also evaluate $M.posBefore(e,L)$ and $M.posAfter(e,L)$ which return the index of the element e in the list of the specified type or the index of the first element $f$ of that type which appears before (or after) that element in the edit-string. This can be useful when we want to compute the first character that appears before a start marker for example. We will show that this can also be computed in time $O(\log(N))$.
